Alone Together: Compositional Reasoning and Inference for Weak Isolation

### 2017-10-18 introduction.tex
Added missing file
edits upto Sec. 5

The tension between serializable transactions, which are easy to
reason about but difficult to implement efficiently, and more
pragmatic variants that are driven by performance and availability
considerations, has motivated the development of weaker forms of
transaction isolation, beginning as early as 1976~\cite{gray1976};
these instantiations permit a transaction to witness various effects
of newly committed, or even concurrently running, transactions while
it executes.  The ANSI SQL 92 standard defines three such weak
isolation levels which are now implemented in many relational and
NoSQL databases. Not surprisingly, weakly-isolated transactions have
been found to significantly outperform serializable transactions on
benchmark suites, both on single-node databases and multi-node
replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
popular ACID and ``NewSQL'' databases found that only three of them
offer serializability by default, and half, including Oracle 11g, do
not offer it at all.  A 2015 study~\cite{bailisferal} of a large
corpus of database applications finds no evidence that applications
manifestly change the default isolation level offered by the
database. Taken together, these studies make clear that
weakly-isolated transactions are quite prevalent in practice, whereas
serializable transactions are often eschewed.

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
  been echoed by leading database researchers who lament the lack of a
  better understanding of this problem.  See: {\tt
    http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

A major problem with weak isolation is that its semantics in the
context of user programs is not precisely understood. The original
proposal~\cite{gray1976} defines multiple ``degrees'' of weak
isolation in terms of implementation details such as the nature and
duration of locks held in each case. The ANSI SQL 92 standard defines
four levels of isolation (including serializability) in terms of
various undesirable \emph{phenomena} (\eg \emph{dirty reads}) each is
required to prevent. While this is an improvement, it requires
programmers to be prescient about the possible ways various
undesirable phenomena might manifest in their applications, and in
each case determine if the phenomenon can be allowed without violating
application invariants. This is understandably hard, especially in the
absence of any formal underpinning to define weak isolation semantics.
Adya~\cite{adyaphd} presents the first formal definitions of some
well-known isolation levels in the context of a sequentially
consistent (SC) database.  However, there has been little progress
relating Adya's system model to a formal operational semantics or a
proof system that can facilitate rigorous correctness arguments.
Consequently, reasoning about weak isolation remains an error prone
endeavor, with major database vendors~\cite{postgresiso, mysqliso,
  oracleiso} continuing to document their isolation levels primarily
in terms of the undesirable phenomena a particular isolation level may
induce, placing the burden on the programmer to determine application
correctness.

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that manifests on replicated data
stores, where atomic operations are concurrently executed against
different replicas, resulting in an execution order inconsistent with
any sequential order. In contrast, weak isolation is a property of
concurrent transactions interfering with one another resulting in an
execution order that is not serializable. Unlike weak consistency,
weak isolation can manifest even in an unreplicated setting, as
evident from the support for weakly-isolated transactions on
conventional (unreplicated) databases as mentioned above.

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support that realizes
this goal.  We develop a set of syntax-directed compositional proof
rules that allow developers to build correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules would likely discourage applications
programmers from using them, we present an inference procedure based
that automatically verifies the weakest isolation level for a
transaction that ensures its consistency requirements are maintained.
The key to inference is a novel formulation of database state
(represented as sets of tuples) as a monad, and in which database
computations are interpreted as state transformers over these sets.
This interpretation leads to an encoding of database computations
amenable for verification by off-the-shelf SMT solvers.  The paper
makes the following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
\item We analyze properties of weak isolation in the context of an
  embedded DSL for OCaml that treats SQL-based relational database
  operations (e.g., inserts, selects, deletes, joins, updates, etc.)
  as monadic computations over an abstract database state.
 \item We develop an operational semantics and compositional proof
   system for this language capable of relating high-level application
   invariants to database state, parameterized by a weak isolation
   semantics that selectively exposes the visibility of these
   operations to other transactions.
 \item Exploiting the monadic definition of database state, we devise
   an inference algorithm capable of discovering the weakest isolation
   level that is sound with respect to a transaction's high-level
   consistency requirements.  The algorithm interprets database
   operations as state transformers expressed in a language amenable
   for translation into a decidable fragment of first-order logic, and
   is thus suitable for automated verification using off-the-shelf SMT
   solvers.
 \item We present details of a realistic implementation along with an
   evaluation study on real-world database benchmarks that justify our
   approach and the utility of our proof methodology and inference
   mechanism..
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

>>>>

Because serializable transactions favor correctness over performance,
there has been long-standing interest~\cite{gray1976} in the database
community to consider weaker variants that try to recover performance,
even at the expense of simplicity and ease of reasoning.  These
instantiations permit a transaction to witness various effects of
newly committed, or even concurrently running, transactions while it
executes, thus weakening serializability's strong isolation
guarantees.  The ANSI SQL 92 standard defines three such weak
isolation levels which are now implemented in many relational and
NoSQL databases. Not surprisingly, weakly-isolated transactions have
been found to significantly outperform serializable transactions on
benchmark suites, both on single-node databases and multi-node
replicated stores~\cite{dbtuningbook,bailishat,bailisvldb}, leading to
their overwhelming adoption. A 2013 study~\cite{bailishotos} of 18
popular ACID and ``NewSQL'' databases found that only three of them
offer serializability by default, and half, including Oracle 11g, do
not offer it at all.  A 2015 study~\cite{bailisferal} of a large
corpus of database applications finds no evidence that applications
manifestly change the default isolation level offered by the
database. Taken together, these studies make clear that
weakly-isolated transactions are quite prevalent in practice, and
serializable transactions are often eschewed.

Unfortunately, weak isolation admits behaviors that are difficult to
comprehend~\cite{berenson}. To quantify weak isolation anomalies,
Fekete \emph{et al.}~\cite{feketevldb09} devised and experimented with
a microbenchmark suite that executes transactions under a
weakly-isolated \emph{read committed} isolation level - the default
level for 8 of the 18 databases studied in~\cite{bailishotos}, and
found that 25 out of every 1000 rows in the database violate at least
one integrity constraint. Bailis \emph{et al.}~\cite{bailisferal} rely
on Rails' \emph{uniqueness validation} to maintain uniqueness of
records while serving Linkbench's~\cite{linkbench} insertion workload
(6400 records distributed over 1000 keys; 64 concurrent clients), and
report discovering more than 10 duplicate records.  Rails relies on
database transactions to validate uniqueness during insertions, which
is sensible if transactions are serializable, but incorrect under the
weak isolation level used in the experiments. The same study has found
that 13\% of all invariants among 67 open source Ruby-on-Rails
applications are at risk of being violated due to weak
isolation. Indeed, incidents of safety violations due to executing
applications in a weakly-isolated environment have been reported on
web services in production~\cite{starbucksbug, scimedbug}, including
in safety-critical applications such as bitcoin
exchanges~\cite{poloniexbug, bitcoinbug}. While enforcing
serializability for all transactions would be sufficient to avoid
these errors and anomalies, it would likely be an overly conservative
strategy; indeed, 75\% of the invariants studied in~\cite{bailisferal}
were shown to be preserved under some form of weak isolation.  When to
use weak isolation, and in what form, is therefore a prominent
question facing all database programmers.\footnote{This position has
been echoed by leading database researchers who lament the lack of a
better understanding of this problem; see e.g., {\tt
  http://www.bailis.org/blog/understanding-weak-isolation-is-a-serious-problem}.}

A major problem with weak isolation as currently specified is that its
semantics in the context of user programs is not easily
understood. The original proposal~\cite{gray1976} defines multiple
``degrees'' of weak isolation in terms of implementation details such
as the nature and duration of locks held in each case. The ANSI SQL 92
standard defines four levels of isolation (including serializability)
in terms of various undesirable \emph{phenomena} (\eg \emph{dirty
  reads} - reading data written by an uncommitted transaction) each is
required to prevent. While this is an improvement, this style of
definition still requires programmers to be prescient about the
possible ways various undesirable phenomena might manifest in their
applications, and in each case determine if the phenomenon can be
allowed without violating application invariants. This is
understandably hard, especially in the absence of any formal
underpinning to define weak isolation semantics.  Adya~\cite{adyaphd}
presents the first formal definitions of some well-known isolation
levels in the context of a sequentially consistent (SC) database.
However, there has been little progress relating Adya's system model
to a formal operational semantics or a proof system that can
facilitate rigorous correctness arguments.  Consequently, reasoning
about weak isolation remains an error prone endeavor, with major
database vendors~\cite{postgresiso, mysqliso, oracleiso} continuing to
document their isolation levels primarily in terms of the undesirable
phenomena a particular isolation level may induce, placing the burden
on the programmer to determine application correctness.

Recent results on reasoning about application invariants in the
presence of weak consistency~\cite{burckhardt14, redblueosdi,
  redblueatc, ecinec, gotsmanpopl16} address broadly related concerns.
Weak consistency is a phenomenon that manifests on replicated data
stores, where atomic operations are concurrently executed against
different replicas, resulting in an execution order inconsistent with
any sequential order. In contrast, weak isolation is a property of
concurrent transactions interfering with one another, resulting in an
execution order that is not serializable. Unlike weak consistency,
weak isolation can manifest even in an unreplicated setting, as
evident from the support for weakly-isolated transactions on
conventional (unreplicated) databases as mentioned above.

@@

In this paper, we propose a program logic for weakly-isolated
transactions along with automated verification support to allow
developers to verify the soundness of their applications, without
having to resort to low-level operational reasoning as they are forced
to do currently.  We develop a set of syntax-directed compositional
proof rules that enable the construction of correctness proofs for
transactional programs in the presence of a weakly-isolated
concurrency control mechanism.  Realizing that the proof burden
imposed by these rules may discourage applications programmers from
using them, we also present an inference procedure based on these
rules that automatically verifies the weakest isolation level for a
transaction that still ensures its invariants are maintained.  The key
to inference is a novel formulation of database state (represented as
sets of tuples) as a monad, and in which database computations are
interpreted as state transformers over these sets.  This
interpretation leads to an encoding of database computations amenable
for verification by off-the-shelf SMT solvers.  The paper makes the
following contributions:
%% We have realized these ideas as an embedded DSL within OCaml, that
%% supports common relational SQL operations (updates, selects, inserts,
%% deletes, joins, etc.).  Experimental results on real-world benchmarks
%% demonstrate the feasibility and value of automated verification for
%% weakly isolated transactions, an important advance in improving the
%% safety of realistic database computations.
\begin{enumerate}
\item We analyze properties of weak isolation in the context of an
    embedded DSL for OCaml that treats SQL-based relational database
    operations (e.g., inserts, selects, deletes, joins, updates, etc.)
    as monadic computations over an abstract database state.
 \item We develop an operational semantics and a compositional
   rely-guarantee style proof system for this language capable of
   relating high-level application invariants to database state,
   parameterized by a weak isolation semantics that selectively
   exposes the visibility of these operations to other transactions.
 \item Exploiting the monadic definition of database state, we devise
   an inference algorithm capable of discovering the weakest isolation
   level that is sound with respect to a transaction's high-level
   consistency requirements.  The algorithm interprets database
   operations as state transformers expressed in a language amenable
   for translation into a decidable fragment of first-order logic, and
   is thus suitable for automated verification using off-the-shelf SMT
   solvers.
 \item We present details of a realistic implementation along with an
   evaluation study on real-world database benchmarks that justify our
   approach and the utility of our proof methodology and inference
   mechanism.
\end{enumerate}
\noindent Our results provide the first (to the best of our knowledge)
formalization of weakly-isolated transactions, along with an
expressive and compositional proof automation framework capable of
verifying the safety of high-level consistency conditions attached to
these transactions.  Collectively, these contributions allow
weakly-isolated transactions to enjoy the same rigorous reasoning
capabilities as their strongly-isolated (serializable) counterparts.

### 2017-10-18 isolation.tex
edits upto Sec. 5

\textbf{Repeatable Read (RR)} By definition, multiple reads to a
transactional variable in a Repeatable Read transaction are required
to return the same value.  RR is often implemented (for e.g., in
~\cite{mysqliso,bailishat}) by executing the transaction against a
(conceptual) snapshot of the database, but committing its writes to
the actual database. This implementation of RR can be axiomatized as
$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
is stronger than the ANSI SQL specification, which requires no more
than the invariance of already read records. In particular, ANSI SQL
RR allows \emph{phantom reads}, a phenonmenon in which a repeated
\C{SELECT} query might return newly inserted records that were not
previously returned. This specification is implemented, for e.g., in
Microsoft's SQL server, using record-level exclusive read locks, that
prevent a record from being modified while it is read by an
uncommitted transaction, but which does not prohibit insertion of new
records. The ANSI SQL RR specification can be axiomatized in our
framework, but it requires a minor extension to our operational
semantics to track a transaction's reads. In particular, the records
returned by \C{SELECT} should be added to the local database $\stl$,
but without changing their transaction identifiers ($\txnf$ fields),
and flush ($\gg$) should only flush the records that bear the current
transaction's identifier. With this extension, ANSI SQL RR can be
axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
      r \in \Delta \Rightarrow r \in \Delta'\\
  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
\end{array}
\end{smathpar}
If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
record written by a different transaction and read by the current
transaction (since the current transaction's records are not yet
present in $\stg$). By requiring $r\in\stg'$, $\I_e$ guarantees the
invariance of $r$, thus the repeatability of the read.

>>>>

\textbf{Repeatable Read (RR)} By definition, multiple reads to a
transactional variable in a Repeatable Read transaction are required
to return the same value.  RR is often implemented (for e.g., in
~\cite{mysqliso,bailishat}) by executing the transaction against a
(conceptual) snapshot of the database, but committing its writes to
the actual database. This implementation of RR can be axiomatized as
$\I_e = \I_{ss}$ and $\I_{c}=true$. However, this specification of RR
is stronger than the ANSI SQL specification, which requires no more
than the invariance of already read records. In particular, ANSI SQL
RR allows \emph{phantom reads}, a phenomenon in which a repeated
\C{SELECT} query might return newly inserted records that were not
previously returned. This specification is implemented, for e.g., in
Microsoft's SQL server, using record-level exclusive read locks, that
prevent a record from being modified while it is read by an
uncommitted transaction, but which does not prohibit insertion of new
records. The ANSI SQL RR specification can be axiomatized in our
framework, but it requires a minor extension to our operational
semantics to track a transaction's reads. In particular, the records
returned by \C{SELECT} should be added to the local database $\stl$,
but without changing their transaction identifiers ($\txnf$ fields),
and flush ($\rhd$) should only flush the records that bear the current
transaction's identifier. With this extension, ANSI SQL RR can be
axiomatized thus:
\begin{smathpar}
\begin{array}{lcl}
  \I_e(\stl,\stg,\stg') & \Leftrightarrow & \forall(r\in\stl).
      r \in \Delta \Rightarrow r \in \Delta'\\
  \I_c(\stl,\stg,\stg') & \Leftrightarrow & true\\
\end{array}
\end{smathpar}
If a record $r$ belongs to both $\stl$ and $\stg$, then it must be a
record written by a different transaction and read by the current
transaction (since the current transaction's records are not yet
present in $\stg$). By requiring $r\in\stg'$, $\I_e$ guarantees the
invariance of $r$, thus the repeatability of the read.
